---
title: "Presentation"
author: "Carson G Miller Rigoli"
date: today
format:
  html:
    theme: pulse
    embed-resources: true
---

## Survey Summary

This report outlines a survey conducted to help understand the preferred class arrival times of UW students.

## Acquire Data

To collect data for this survey, I used a convenience sample of 6 students that sat near me in my INFO class in Autumn 2024. I asked each for their name and for when they ideally would prefer to arrive to class. I specifically asked how many seconds before (negative) or after (positive) the beginning of class they would prefer to be seated.

Here is the data I collected:

```{r}
#| echo: false

# Enter Measurements
names <- c("Maria", "Linh", "Ravi", "Esther", "Lee")
preferred_times <- c(0, -20, -60, -120, -1200)

# Convert into a data frame and store with a name
survey_data <- data.frame(name = names, preferred_time = preferred_times)

# Show data frame
survey_data
```
## Tranform Data

For this analysis, I want to report times in minutes rather than seconds, so I created a new variable to record preferred times in minutes by converting the time preferred time variable.

```{r}
#| echo: false

survey_data$preferred_time_min <- survey_data$preferred_time / 60

```

## Summarize Data

I calculated summary statistics to give me a sense of the data.

```{r}
summary(survey_data)
```
The average preferred arrival time was around 4.7 minutes before the class start.

## Interpret Findings

Based on the survey data, my best guess for the average preferred arrival time for UW students is 4.7 minutes before class starts. While this might be approximately correct, I do not have a lot of confidence in these results.

This study had a number of limitations:

* I only collected a very small sample of data
* I only collected data from students in an introductory informatics course
* I only collected data from students who had already arrived in a class


